---
title: "Introduction to BivLaplaceRL"
author: "Mahesh Divakaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BivLaplaceRL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
uni_fns_available <- tryCatch(
  exists("lt_residual", asNamespace("BivLaplaceRL")),
  error = function(e) FALSE
)
```

## Overview

**BivLaplaceRL** is an R package that implements statistical methods from two
research papers on bivariate Laplace transforms in reliability theory, together
with a set of univariate residual life tools:

1. Bivariate Laplace transform of residual lives (Jayalekshmi, Rajesh & Nair, 2022)
2. Bivariate Laplace transform order of reversed residual lives (Jayalekshmi & Rajesh)

```{r setup}
library(BivLaplaceRL)
```

## Parametric Distributions

The package provides four bivariate distributions commonly used in reliability
modelling.

### Gumbel Bivariate Exponential

```{r gumbel}
set.seed(42)
dat <- rgumbel_biv(200, k1 = 1, k2 = 1.5, theta = 0.3)
head(dat)
sgumbel_biv(1, 1, k1 = 1, k2 = 1.5, theta = 0.3)
```

### Farlie-Gumbel-Morgenstern (FGM)

```{r fgm}
set.seed(1)
dat_fgm <- rfgm_biv(100, theta = 0.5)
pfgm_biv(0.5, 0.6, theta = 0.5)
```

### Schur-Constant

```{r schur}
set.seed(2)
dat_sc <- rschur_biv(100, lambda = 1)
summary(rowSums(dat_sc))
```

## Bivariate Laplace Transform of Residual Lives

### Closed-Form (Gumbel Distribution)

$$L^*_{X_{t_1|t_2}}(s_1) = \frac{1}{k_1 + s_1 + \theta t_2}, \quad
  L^*_{X_{t_2|t_1}}(s_2) = \frac{1}{k_2 + s_2 + \theta t_1}.$$

```{r closed_form}
blt_residual_gumbel(s1 = 1, s2 = 1, t1 = 0.5, t2 = 0.5,
                    k1 = 1, k2 = 1, theta = 0.5)
```

### Numerical Computation

```{r numerical}
my_surv <- function(x1, x2) exp(-x1 - x2 - 0.3 * x1 * x2)
blt_residual(s1 = 1, s2 = 1, t1 = 0.5, t2 = 0.5,
             surv_fn = my_surv, star = TRUE)
```

### Nonparametric Estimation

```{r np_estim}
set.seed(123)
dat <- rgumbel_biv(500, k1 = 1, k2 = 1, theta = 0.5)
np_blt_residual(dat, s1 = 1, s2 = 1, t1 = 0.3, t2 = 0.3)

# True value
blt_residual_gumbel(s1 = 1, s2 = 1, t1 = 0.3, t2 = 0.3, theta = 0.5)
```

### Simulation Study

```{r sim_study}
sim_blt_residual(n_obs = 200, n_sim = 50, s1 = 1, s2 = 1,
                 t1 = 0.3, t2 = 0.3, k1 = 1, k2 = 1, theta = 0.5)
```

## Bivariate Hazard Gradient and MRL

```{r hazard_mrl}
biv_hazard_gradient(t1 = 1, t2 = 1, k1 = 1, k2 = 1, theta = 0.3)
biv_mean_residual(t1 = 0.5, t2 = 0.5, k1 = 1, k2 = 1, theta = 0)
```

## NBUHR / NWUHR Aging Classification

```{r aging}
res <- nbuhr_test(t2 = 0.5, k1 = 1, k2 = 1, theta = 0.3,
                  t1_grid = seq(0.1, 3, 0.3))
cat("Component 1:", res$class1, "\n")
cat("Component 2:", res$class2, "\n")
```

## Bivariate Laplace Transform of Reversed Residual Lives

```{r reversed}
blt_reversed(s1 = 1, s2 = 1, t1 = 0.5, t2 = 0.5, theta = 0.3)
biv_rhazard_gradient(x1 = 0.5, x2 = 0.5, theta = 0.3)
biv_rmrl(x1 = 0.5, x2 = 0.5, theta = 0.3)
```

## Bivariate Stochastic Orders

```{r blt_order}
sX <- function(x1, x2) sgumbel_biv(x1, x2, k1 = 1, k2 = 1, theta = 0.2)
sY <- function(x1, x2) sgumbel_biv(x1, x2, k1 = 2, k2 = 2, theta = 0.2)
res <- blt_order_residual(sX, sY, s1 = 1, s2 = 1,
                          t1_grid = c(0.5, 1), t2_grid = c(0.5, 1))
cat("X <=_BLt-rl Y:", res$order_holds, "\n")

res2 <- biv_whr_order(sX, sY, t1_grid = c(0.5, 1), t2_grid = c(0.5, 1))
cat("X <=_whr Y:", res2$order_holds, "\n")
```

## Univariate Residual Life Analysis

The package now provides a complete set of univariate residual life tools
that complement the bivariate framework.

### Laplace Transform of Residual Life

For a non-negative continuous random variable $X$ with density $f$ and
survival function $\bar{F}$:

$$L_X(s,t) = E[e^{-sX} \mid X > t]
  = \frac{1}{\bar{F}(t)}\int_t^\infty e^{-sx} f(x)\,dx.$$

```{r uni_lt, eval = uni_fns_available}
f  <- function(x) dexp(x, 1)
Fb <- function(x) pexp(x, 1, lower.tail = FALSE)

# At t=0: L_X(1, 0) = 1/(1+1) = 0.5
lt_residual(f, Fb, s = 1, t = 0)

# At t=0.5
lt_residual(f, Fb, s = 1, t = 0.5)
```

### Hazard Rate and MRL

```{r uni_hr_mrl, eval = uni_fns_available}
# Exp(1): constant hazard rate = 1
hazard_rate(f, Fb, t = c(0.5, 1, 2))

# Exp(1): constant MRL = 1 (memoryless)
mean_residual(Fb, t = c(0, 0.5, 1, 2))

# Gamma(2,1): increasing hazard, decreasing MRL
fG  <- function(x) dgamma(x, shape = 2, rate = 1)
FbG <- function(x) pgamma(x, shape = 2, rate = 1, lower.tail = FALSE)
hazard_rate(fG, FbG, t = c(0.5, 1, 2))
mean_residual(FbG, t = c(0, 0.5, 1))
```

### Nonparametric Estimation

```{r uni_np, eval = uni_fns_available}
set.seed(7)
x <- rexp(500, rate = 1)

# Nonparametric estimate at s=1, t=0: true value = 0.5
np_lt_residual(x, s = 1, t = 0)

# At t=0.5
np_lt_residual(x, s = 1, t = 0.5)
```

### Univariate Stochastic Orders

```{r uni_orders, eval = uni_fns_available}
f1  <- function(x) dexp(x, 1)
Fb1 <- function(x) pexp(x, 1, lower.tail = FALSE)
f2  <- function(x) dexp(x, 2)
Fb2 <- function(x) pexp(x, 2, lower.tail = FALSE)

# LT-rl order: Exp(1) <=_Lt-rl Exp(2)?
lt_rl_order(f1, Fb1, f2, Fb2,
            s_grid = c(0.5, 1, 2), t_grid = c(0, 0.5, 1))$order_holds

# Hazard rate order: Exp(1) <=_hr Exp(2)?
hr_order(f1, Fb1, f2, Fb2, t_grid = c(0.5, 1, 2))$order_holds

# MRL order: Exp(2) <=_mrl Exp(1)?
mrl_order(Fb2, Fb1, t_grid = c(0, 0.5, 1, 2))$order_holds
```

## Entropy and Information Generating Functions

```{r entropy}
# Shannon entropy of Exp(1) = 1
shannon_entropy(function(x) dexp(x, 1))

# Golomb IGF of Exp(1) at alpha=2: 0.5
info_gen_function(function(x) dexp(x, 1), alpha = 2)
```

## References

Jayalekshmi S., Rajesh G., Nair N.U. (2022). Bivariate Laplace transform of
residual lives and their properties. *Communications in Statistics---Theory and
Methods*. <https://doi.org/10.1080/03610926.2022.2085874>

Jayalekshmi S., Rajesh G., Nair N.U. (2022). Bivariate Laplace transform order
and ordering of reversed residual lives. *International Journal of Reliability,
Quality and Safety Engineering*. <https://doi.org/10.1142/S0218539322500061>

Belzunce F., Ortega E., Ruiz J.M. (1999). The Laplace order and ordering of
residual lives. *Statistics & Probability Letters*, 42(2), 145--156.
<https://doi.org/10.1016/S0167-7152(98)00202-8>
